---
title: "IHME Technical Assessment"
author: "Hans Elliott"
date: "5/5/2022"
output: 
  html_document:
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```


```{r warning=FALSE, message=FALSE, results='hide'}
## DATA & LIBRARIES ------------------------

# Load libraries
## Install the 'pacman' package-management package if necessary 
if (!require("pacman")) install.packages("pacman")
## Install other packages using pacman::p_load()
pacman::p_load(here, janitor, skimr,        ##for loading/cleaning data
               ggplot2, RColorBrewer, ggrepel, scales,  ##for plotting
               forecast,    ##for some forecasting
               Hmisc,       ##for creating a lag variable
               dplyr, tidyr ##for general data manipulation
               )

# Load Data
covid = here("data", "covid_data_cases_deaths_hosp.csv") %>% read.csv()
dim(covid) ##view data dimensions

## CLEANING ---------------------------------
## Convert variable names to common format ("snake_case")
covid = janitor::clean_names(covid)
names(covid)

## View # of missing values in each variable
# (Hospitalization data is available for only some states)
skimr::skim(covid) %>% select(skim_variable, n_missing, complete_rate)

```



## Q1: Cases, Hospitalizations, & Deaths
COVID-19 cases, hospitalizations, and deaths share an almost linear relationship. Plot 1 shows the relationship between these indicators grouped by state and totaled over time (2020). The states with the lowest reported cases also had the least deaths and the least hospitalizations. As we move along the x-axis, higher levels of reported cases map to states with higher deaths and hospitalizations. This is logical since states which suffered more severe outbreaks had more COVID cases, yielding more hospitalizations and more deaths.     
Plot 2 shows the accumulation of cases, deaths, and hospitalizations over time. Although there are significantly more reported cases than deaths or hospitalizations, the three indicators follow very similar time trends, providing further evidence of a linear relationship with each other. However, Plot 2 also reveals that these variables are not linear in time. The first several months of 2020 saw sharp increases in all three variables which began to level out towards the second half of the year.  




#### Plot 1: Relationship by Location
```{r, dpi = 200, warning=FALSE,message=FALSE}
## Create color palette for plot
my_palette = colorRampPalette(RColorBrewer::brewer.pal(9, "YlOrRd"))

##Summarize COVID indicators by location
covid_by_place = covid %>% na.omit() %>% ##omit observations with missing values
  ##Summarize by location
  group_by(location_id) %>%
  summarise(confirmed = sum(confirmed, na.rm = T),
            deaths = sum(deaths, na.rm = T),
            hosp = sum(hospitalizations, na.rm = T),
            state = province_state,
            population = population) %>% unique()

##Create plot
covid_by_place %>%  
  ggplot(aes(x = confirmed/1e3, y = deaths/1e3)) +
    geom_point(alpha = 0.7, 
               aes(color = hosp/1e3, size = hosp/1e3)) +
    ##Set color and size scale to combine those legends
    scale_color_gradientn(colors = my_palette(100),
                          limits=c(0, 1000), breaks=seq(0, 1000, by=250)) +
    scale_size_continuous(limits=c(0, 1000), breaks=seq(0, 1000, by=250)) +
    guides(color = guide_legend(), size = guide_legend()) +
    ##Set axis scales
    scale_x_continuous(breaks = seq(0, 6000, by = 1000),limits = c(0, 6000)) +
    scale_y_continuous(breaks = seq(0, 300, by = 50)) +
    ##Add location labels
    ggrepel::geom_label_repel(data = covid_by_place %>% ##filter to extreme
                                filter(confirmed/1e3 > 4000),
                              aes(label = state),
                              size = 2, box.padding = unit(0, "lines"),
                              force = 50, alpha = 0.5
                              ) +
    ##Add plot labels
    labs(
      title = "COVID-19 by State: Cumulative Cases, Deaths, & Hospitalizations",
      subtitle = "Indicators totaled across time",
      x = "Thousands of Reported Cases",
      y = "Thousands of Deaths", 
      color = "Hospitalizations \n (Thousands)",
      size = "Hospitalizations \n (Thousands)",
      caption = "Sample of locations for which hospitalizations are reported.") +
    ##create aesthetic theme
    theme_minimal() + 
    theme(legend.position = "right",
          legend.direction = "vertical",
          plot.title = element_text(size = "11", family = "sans"),
          plot.subtitle = element_text(size="9",family="sans",color="gray45"),
          axis.title = element_text(size="9",family="sans",color="gray45"),
          axis.text = element_text(size="9",family="sans",color="gray45"),
          legend.title = element_text(size="9",family="sans",color="gray45"),
          legend.text = element_text(size="8",family="sans",color="gray45"),
          plot.caption = element_text(size="6",family="sans",color="gray45"),
          panel.grid.major = element_line(color = "gray93"),
          panel.border = element_blank(),
          panel.background = element_blank()
          #panel.grid.minor = element_line(color = "gray93")
              )

```


#### Plot 2: Relationship Over Time

```{r, dpi=200, results='asis'}

##Total COVID indicators for each date
covid_by_date = covid %>%
  group_by(date) %>%
  summarise(cases = sum(confirmed, na.rm = T),
            deaths = sum(deaths, na.rm = T),
            hosp = sum(hospitalizations, na.rm = T)) %>%
  ##Convert date from EXCEL's numeric format to date format
  mutate(date = as.Date(date, origin = "1899-12-30"))


##Mutate vars for plotting
covid_by_date %>%
  transmute(
    'Log Cases' = ifelse(cases==0, 0, log(cases)),
    'Log Deaths' = ifelse(deaths==0, 0, log(deaths)),
    'Log Hospitalizations' = ifelse(hosp==0, 0, log(hosp)),
    date = date
  ) %>%
  ##Pivot data longer for facet wrapping
  pivot_longer(cols = !date,
               names_to = "indicator", 
               values_to = "count") %>%
  
  
##Create plot  
  ggplot(aes(x = date, y = count, 
             color = indicator, fill = indicator)) +
    geom_line(alpha = 0.7) +
    geom_area(alpha = 0.5) +
    facet_wrap(~indicator, scales = "free") +
    ##format x axis (date axis)
    scale_x_date(date_breaks = "1 month",
                 labels = scales::date_format("%b")) +
    ##add plot labels
    labs(title = "U.S. 2020 Covid-19 Outbreak: Changes over Time",
         x = "2020", y = "Log of Cumulative Count (Thousands)",
         caption = 
           "Hospitalization data available for a limited number of states.") +
    ##format theme
    theme_minimal() +
    theme(legend.position = "none",
          plot.title = element_text(size = "11", family = "sans"),
          plot.subtitle = element_text(size="9",family="sans",color="gray45"),
          axis.title = element_text(size="9",family="sans",color="gray45"),
          axis.text = element_text(size="9",family="sans",color="gray45"),
          legend.title = element_text(size="9",family="sans",color="gray45"),
          legend.text = element_text(size="8",family="sans",color="gray45"),
          plot.caption = element_text(size="6",family="sans",color="gray45"),
          panel.grid.major = element_line(color = "gray93"),
          panel.border = element_blank(),
          panel.background = element_blank())

```


## Q2: Modeling Daily Deaths

I fit a linear regression to predict daily U.S. COVID-19 deaths from cases and hospitalizations. Before fitting the model, I aggregated the cases, hospitalizations, and deaths by date, providing a total across the U.S. for each day. I then split this dataset into a training and testing (validation) set in order to evaluate different model specifications. I settled on a model which predicts daily deaths given cases, hospitalizations, a 14-day lag variable of deaths, and a death to case ratio which is also lagged by 14 days. This was one of the less flexible specifications that I tried, which will ideally limit overfitting. Ultimately it was the best model in terms of reducing training and validation RMSE and still providing reasonable projections. As seen in the plot below, it is effective at capturing the general trend in COVID-19 deaths over time.  
```{r, echo=FALSE, results='hide'}
# DATA PREPROCESSING ------------------------------------------------

##Aggregate variables by date across the US
model_df = covid %>% 
  group_by(date) %>%
  summarise(cases = sum(confirmed, na.rm = T),
            deaths = sum(deaths, na.rm = T),
            hosp = sum(hospitalizations, na.rm = T),
            hosp_case_ratio = hosp/cases) %>%
 ##Convert date from EXCEL's numeric format to date format 
  mutate(date = as.Date(date, origin = "1899-12-30"),
        #& add 14-day lag var for deaths and cases
        #need to correct for the NAs introduced at beginning of series
         lagged_deaths = Hmisc::Lag(deaths, +14),
          lagged_deaths = ifelse(is.na(lagged_deaths), 0, lagged_deaths),
         lagged_cases = Hmisc::Lag(cases, +14),
          lagged_cases = ifelse(is.na(lagged_cases), 0, lagged_cases),
         lag_case_death_rat = lagged_deaths/lagged_cases,
          lag_case_death_rat = ifelse(is.na(lag_case_death_rat), 0,
                                      lag_case_death_rat)) %>%
 ##Center/rescale covariates to avoid overflow errors
 ##(was necessary for inclusion of polynomials or interaction terms)
  mutate(scaled_cases = scale(cases),
         scaled_hosp = scale(hosp))


##Split data into training and testing (before and after June 1)
train_df = model_df %>% filter(date < "2020-06-01") 
test_df = model_df %>% filter(date >= "2020-06-01")

## TRAINING FIT ---------------------------------------------

mod1 = lm(deaths ~ lagged_deaths + cases + lag_case_death_rat,
                     data = train_df)

summary(mod1) ##view model summary

## View model RMSE
print(paste("Model Training RMSE:",
            sqrt( mean((train_df$deaths - predict(mod1))^2)  ) %>% round(4)
))

#Model Testing ------------------------------------------------

## VIEW TRAINING PREDICTIONS
##predicted deaths
training_pred = data.frame(date = train_df$date,
                        death_hat = predict(mod1))

ptrain = ggplot() +
  geom_line(data = train_df, aes(x = date, y = deaths), 
            size = 1, alpha = 0.5) +
  geom_line(data = training_pred,
            aes(x = date, y = death_hat),
            size = 0.5, alpha = 0.9, color = 'red')

## VIEW TESTING PREDICTIONS
test_pred = data.frame(death_hat = predict(mod1, test_df),
                        date = test_df$date)

print(paste("Model Testing RMSE:",
            sqrt( mean((test_df$deaths - test_pred$death_hat)^2) ) %>%
              round(5)
))

ptest = ggplot() +
  geom_line(data = test_df, aes(x = date, y = deaths), 
            size = 1, alpha = 0.5) +
  geom_line(data = test_pred,
            aes(x = date, y = death_hat),
            size = 0.5, alpha = 0.9, color = 'red')
```


```{r, echo=FALSE, dpi=200}
##FINAL MODEL -------------------------------------------
## Fit final model on full dataset
mod_final = lm(deaths ~ cases + hosp + lagged_deaths + lag_case_death_rat,
               data = model_df)

## Extract predicted deaths
mod_pred = data.frame(date = model_df$date,
                      death_hat = predict(mod_final))

## Plot results
ggplot() +
  geom_line(data = model_df, aes(x = date, y = deaths/1e3), 
            size = 1, alpha = 0.5) +
  geom_line(data = mod_pred, aes(x = date, y = death_hat/1e3),
            size = 0.5, alpha = 0.9, color = 'red') +
  ##format x axis (date axis)
  scale_x_date(date_breaks = "1 month",
                 labels = scales::date_format("%b")) +
  ##add line labels
  annotate("text", x = as.Date("2020-05-01", origin = "2020-05-01"),
           y = 125, label = "True Deaths", 
           size = 3.5, alpha = 0.5) +
  annotate("text", x = as.Date("2020-05-01", origin = "2020-05-01"),
           y = 115, label = "Predicted Deaths", 
           color = "red", size = 3.5, alpha = 0.9) +
  ##add plot labels
  labs(title = "Actual and Predicted Cumulative COVID-19 Deaths (2020)",
       x = "2020", y = "Thousands of Deaths (Daily)") +
  ##format theme
    theme_minimal() +
    theme(legend.position = "none",
          plot.title = element_text(size = "11", family = "sans"),
          plot.subtitle = element_text(size="9",family="sans",color="gray45"),
          axis.title = element_text(size="9",family="sans",color="gray45"),
          axis.text = element_text(size="9",family="sans",color="gray45"),
          legend.title = element_text(size="9",family="sans",color="gray45"),
          legend.text = element_text(size="8",family="sans",color="gray45"),
          plot.caption = element_text(size="6",family="sans",color="gray45"),
          panel.grid.major = element_line(color = "gray93"),
          panel.border = element_blank(),
          panel.background = element_blank())

```



```{r, results='asis'}
p_load(stargazer)
stargazer(mod_final, type = "html",
          covariate.labels = c("Cases", "Hospitializations", 
                               "Deaths (14 day lag)",
                               "Deaths/Cases (14 day lag)"),
          dep.var.caption = "", keep.stat = c("n","rsq"),
          dep.var.labels = "Cumulative COVID-19 Deaths")

```


# Q3: 14-Day Projection  

My projection suggests that cumulative COVID-19 deaths would level out over those next 2 weeks, increasing by only about 5,000 deaths. This result is believable given the trends in the true data. One benefit of this simple linear regression approach is that the model is not too flexible, which should help prevent against overfitting, but may increase bias. Additionally, linear regression is fairly interpretable. I included the model parameters in the table above to show the relative contribution of each variable to predicted deaths. Including the lag variables was crucial for creating projections that were consistent with the observed trend in the data, especially considering the oddly low hospitalization numbers reported in the last few days of the data. Alternatively, the model is not well equipped to predict seasonal-like trends or shocks in COVID deaths, since it assumes a linear relationship between deaths and predictors.  


```{r, dpi=200}
## Preprocessing ----------------------------------
##Forecasting dates:
last_date = max(model_df$date) ##last observed date in the data
project_dates = as.Date(last_date + c(1:14)) ##add 2 weeks


##Forecast predictors to use in projection.
##(I use a random walk forecast with a drift, from the 'forecast' package)
case_pred = forecast::rwf(model_df$cases, h = 14, drift = TRUE) %>%
                data.frame()
hosp_pred = forecast::rwf(model_df$hosp, h = 14, drift = TRUE) %>%
                data.frame()

##Combine into a dataframe
project_df = data.frame(date = project_dates,
                        cases = case_pred$Point.Forecast,
                        hosp = hosp_pred$Point.Forecast) %>%
              mutate(hosp_case_ratio = hosp/cases,
                      ##center/rescale covariates again
                      scaled_cases = scale(cases),
                      scaled_hosp = scale(hosp))
##Add lagged deaths & lagged cases
##First filter to dates of the last 14 observations, and extract. Then reverse
##the order so that the first projected observation is matched with the deaths
##corresponding to 14 days beforehand.
##I subtract 0:13 so that the first lagged date is just the last recorded date.
lagged_deaths = model_df %>% ##filter to last 14 days
        filter(date %in% as.Date(last_date - c(0:13))) %>% 
        select(lagged_deaths)
lagged_cases = model_df %>%
        filter(date %in% as.Date(last_date - c(0:13))) %>% 
        select(lagged_cases)
project_df = cbind(project_df, 
                   lagged_deaths = rev(lagged_deaths),
                   lagged_cases = rev(lagged_cases)) %>%
  mutate(lag_case_death_rat = lagged_deaths/lagged_cases)

##Projection -----------------------------------------
##Project deaths with our model (and add to df)
deaths_proj = predict(mod_final, project_df, interval = "confidence")
project_df = cbind(project_df, deaths_proj)

##Plot ---------------------------------------------
ggplot() +
  #actual deaths
  geom_line(data = model_df, aes(x = date, y = deaths/1e3), 
            size = 1, alpha = 0.5) +
  #projected deaths (confidence interval + line)
  geom_ribbon(data = project_df, aes(x = date, 
                                     ymin = lwr/1e3, ymax = upr/1e3),
              fill = 'red', alpha = 0.3) +
  geom_line(data = project_df,
            mapping = aes(x = date, y = fit/1e3),
            size = 0.5, alpha = 0.9, color = 'red') +
  #modeled deaths
  geom_line(data = mod_pred, aes(x = date, y = death_hat/1e3),
            size = 0.5, alpha = 0.3, color = 'red') +
  ##add line labels
  annotate("text", x = as.Date("2020-05-01", origin = "2020-05-01"),
           y = 125, label = "True Deaths", 
           size = 3, alpha = 0.5) +
  annotate("text", x = as.Date("2020-05-01", origin = "2020-05-01"),
           y = 115, label = "Predicted Deaths", 
           color = "red", size = 3, alpha = 0.5) +
  annotate("text", x = as.Date("2020-08-01", origin = "2020-05-01"),
           y = 115, label = "Projected Deaths \n (95% CI)", 
           color = "red", size = 3.5) +
  ##add plot labels
  labs(title = "COVID-19 Cumulative Deaths: 14-Day Projection",
       x = "2020", y = "Thousands of Deaths") +
  ##format x axis (date axis)
  scale_x_date(date_breaks = "1 month",
                 labels = scales::date_format("%b")) +
  ##format theme
    theme_minimal() +
    theme(legend.position = "none",
          plot.title = element_text(size = "11", family = "sans"),
          plot.subtitle = element_text(size="9",family="sans",color="gray45"),
          axis.title = element_text(size="9",family="sans",color="gray45"),
          axis.text = element_text(size="9",family="sans",color="gray45"),
          legend.title = element_text(size="9",family="sans",color="gray45"),
          legend.text = element_text(size="8",family="sans",color="gray45"),
          plot.caption = element_text(size="6",family="sans",color="gray45"),
          panel.grid.major = element_line(color = "gray93"),
          panel.border = element_blank(),
          panel.background = element_blank())
  
```



## Q4: Future Improvement
While the projections from this model appear somewhat reliable, there is much to be improved upon. Given more time, I would experiment with some classic time series models. For example, an ARIMA model might be a more sophisticated way to predict future deaths using past deaths as a predictor (the "auto-regressive" component) while also incorporating seasonality or some random processes. I would also be interested in the performance of non-parametric methods, like machine learning. While this may be unnecessary and we would probably need more predictors, something like a random forest or boosted ensemble could be implemented quickly and might have decent results.    



